Initial estuarine response to the nutrient dense Piney Point wastewater discharge into Tampa Bay, Florida

### 2021-10-08 manu-draft.Rmd
add Pinellas County to line 335
Pin Co removed the majority of the fish, even in St. Pete
scriveners edits line 345

Fish kill reports attributed to red tide at the cities of Tampa and Saint Petersburg, FL somewhat tracked the annual *K. brevis* cell concentrations, with a notable exception in 2021 which had the most reports among all years (Figure \@ref(fig:redtide)c). Historically, more fish kills were reported in Saint Petersburg (`r annstpper`%) as compared to Tampa (`r anntmpper`%), where the former is closer to the mouth of Tampa Bay.  In 2021, `r wkstpcnt` reports were made in Saint Petersburg and `r wktmpcnt` in Tampa (Figure \@ref(fig:redtide)d).  The combined weekly reports in 2021 for Tampa and Saint Petersburg peaked the week of July 4th, the same week as the peak of *K. brevis* cell concentrations (Figure \@ref(fig:redtide)b).  Notably, the first to last week of fish kill reports covered only one and half months, whereas red tide in the bay was observed for nearly four months.  Increased reports in early July coincided with a shift in winds from Tropical Storm Elsa which moved dead fish closer to populated nearshore areas, as noted earlier.  As a result, the city of St. Petersburg removed over 1800 tons of dead fish near public and private shoreline areas (K. Hammer Levy, Pinellas County, pers. comm. Aug. 2021).

@@

From a seagrass perspective, both major bloom events contributed to a reduction in water clarity and less light availability during the Summer 2021 in comparison to prior years. Future effects on seagrass persistence in this region of Tampa Bay are unknown at this time, but monitoring continues.  Changes in frequency occurrence of seagrasses and associations with water quality and macroalgae were ambiguous and not readily apaprent over the initial 4 month sampling period.  Evaluations of frequency occurrence across months showed no changes, although correlations with water quality and macroalgae can be explained with some certainty.  Spatially, transects closer to the mouth of Tampa Bay and extending to the northern portion of Sarasota Bay (area 3) had higher frequency occurrence of turtle grass and lower frequency of shoal grass.  This reflects a natural partitioning of the species along a salinity gradient [Figure \@ref(fig:allpca), @Sherwood17].  Correlations between macroalgae and seagrasses can also be explained by spatial differences along a depth gradient. Frequency occurrences of both shoal grass and turtle grass were negatively associated with manatee grass.  Shoal grass and turtle grass generally grow at more shallow depths than manatee grass and are often exposed during low tides.  As an adaptive mechanism to varying water levels at shallower depths, blades of shoal grass and turtle grass can "lay down" during low tide to minimize desiccation stress.  Manatee grass has thicker, cylindrical blades that are more rigid and, as a result, the species grows deeper. In addition to species depth partitioning, these associations may also reflect natural variation in seagrass response to water clarity gradients along the main axis of Tampa Bay.  Shoal grass can occur in more turbid waters of oligohaline areas.

>>>>

Fish kill reports attributed to red tide at the cities of Tampa and Saint Petersburg, FL somewhat tracked the annual *K. brevis* cell concentrations, with a notable exception in 2021 which had the most reports among all years (Figure \@ref(fig:redtide)c). Historically, more fish kills were reported in Saint Petersburg (`r annstpper`%) as compared to Tampa (`r anntmpper`%), where the former is closer to the mouth of Tampa Bay.  In 2021, `r wkstpcnt` reports were made in Saint Petersburg and `r wktmpcnt` in Tampa (Figure \@ref(fig:redtide)d).  The combined weekly reports in 2021 for Tampa and Saint Petersburg peaked the week of July 4th, the same week as the peak of *K. brevis* cell concentrations (Figure \@ref(fig:redtide)b).  Notably, the first to last week of fish kill reports covered only one and half months, whereas red tide in the bay was observed for nearly four months.  Increased reports in early July coincided with a shift in winds from Tropical Storm Elsa which moved dead fish closer to populated nearshore areas, as noted earlier.  As a result, Pinellas County and the city of St. Petersburg removed over 1800 tons of dead fish near public and private shoreline areas (K. Hammer Levy, Pinellas County, pers. comm. Aug. 2021).

@@

From a seagrass perspective, both major bloom events contributed to a reduction in water clarity and less light availability during the summer 2021 in comparison to prior years. Future effects on seagrass persistence in this region of Tampa Bay are unknown at this time, but monitoring continues.  Changes in frequency occurrence of seagrasses and associations with water quality and macroalgae were ambiguous and not readily apparent over the initial 4 month sampling period.  Evaluations of frequency occurrence across months showed no changes, although correlations with water quality and macroalgae can be explained with some certainty.  Spatially, transects closer to the mouth of Tampa Bay and extending to the northern portion of Sarasota Bay (area 3) had higher frequency occurrence of turtle grass and lower frequency of shoal grass.  This reflects a natural partitioning of the species along a salinity gradient [Figure \@ref(fig:allpca), @Sherwood17].  Correlations between macroalgae and seagrasses can also be explained by spatial differences along a depth gradient. Frequency occurrences of both shoal grass and turtle grass were negatively associated with manatee grass.  Shoal grass and turtle grass generally grow at more shallow depths than manatee grass and are often exposed during low tides.  As an adaptive mechanism to varying water levels at shallower depths, blades of shoal grass and turtle grass can "lay down" during low tide to minimize desiccation stress.  Manatee grass has thicker, cylindrical blades that are more rigid and, as a result, the species grows deeper. In addition to species depth partitioning, these associations may also reflect natural variation in seagrass response to water clarity gradients along the main axis of Tampa Bay.  Shoal grass can occur in more turbid waters of oligohaline areas.
